DATA 621---Business Analytics and Data Mining

### 2020-09-14 HW1/HW1_draft_kellogg.Rmd
Rough Draft of Final submission
Combined most of the RMD files available here.
Exception HW1_BP.rmd - If we want to add the model here or the ggplot subplot, let me know

---
title: 'DATA 621---Business Analytics and Data Mining'
subtitle: 'Fall 2020---Group 2---Homework #1'
author: Avraham Adler, Samantha Deokinanan, Amber Ferger, John Kellogg,
    Bryan Persaud, Jeff Shamp
date: "9/27/2020"
output:
  pdf_document: default
  html_document: default
---

@@

```{r loadData, include=FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(ggplot2)
library(corrplot)
library(data.table)
train_set <- read.csv('data/moneyball-training-data.csv')
eval_set <- read.csv('data/moneyball-evaluation-data.csv')
```

@@

We are given a data set of 2,276 records containing 15 seasonal statistics and the
total number of wins a team had in a given year. On average, about 50% of games 
played are won (81 games out of 162), with the best season having 146 wins and the worst 
season having 0 wins. The data is normally distributed and most years have between 49 
and 112 wins (blue lines below). The nature of the distribution means there aren't too 
many extreme seasons where wins are significantly higher or lower than usual. This serves 
as a good gut-check for our final predictions; if the predicted wins are too high or too 
low, we know something in our model is probably off.

@@

**Stats with an expected negative impact: ** Intuitively, we expect that Caught 
stealing, Errors, Hits allowed, Homeruns allowed, Strikeouts by batters, and Walks 
allowed would all have a **negative** impact on the total wins. In other words, as 
these values increase, we expect that the team is less likely to win.

@@

negSet <- train_set %>%
  select(TARGET_WINS, 
         'Errors' = TEAM_FIELDING_E, 
         'Hits allowed' = TEAM_PITCHING_H, 
         'Strikeouts by batters' = TEAM_BATTING_SO, 
         'Caught stealing' = TEAM_BASERUN_CS, 
         'Walks allowed' = TEAM_PITCHING_BB, 
         'Homeruns allowed' = TEAM_PITCHING_HR)

@@

Even more interestingly, Homeruns allowed and Walks allowed have the *opposite* 
effect; as these stats increase, so do the number of wins!

**Stats with an expected positive impact: ** We can look at the same information for the 
stats that we expect to have a **positive** effect on wins: Base hits, Doubles, Triples, 
Homeruns, Walks, Batters getting hit by pitches, Stolen bases, Double Plays, and Strikeouts 
by pitchers.

@@

posSet <- train_set %>%
  select(TARGET_WINS, 
         'Base hits' = TEAM_BATTING_H, 
         'Doubles' = TEAM_BATTING_2B, 
         'Triples' = TEAM_BATTING_3B, 
         'Homeruns' = TEAM_BATTING_HR, 
         'Walks' = TEAM_BATTING_BB, 
         'Batters hit' = TEAM_BATTING_HBP,
         'Stolen bases' = TEAM_BASERUN_SB,
         'Double plays' = TEAM_FIELDING_DP,
         'Strikeouts' = TEAM_PITCHING_SO)

@@

Many of these stats *do* seem to have an effect on the number of wins, most notably,
base hits and walks. We see weaker positive relationships for homeruns, doubles, triples, 
and stolen bases. This makes sense when we think about it; these things tend to happen 
less often in games than pure base hits and walks, so they don't have as much of an 
effect on winning. Finally, double plays and batters hit don't appear to have any 
correlation with the number of wins. Once again, this intuitively makes sense because 
they are less likely to happen in a game.

@@

Some of the provided statistics are well-balanced in the sense that there are
very few (or no) extreme values. **Homeruns by batters**, **Strikeouts by** 
**batters**, and **Homeruns allowed** are examples of this.

@@

set1 <-  train_set %>%
  select('Homeruns by batters' = TEAM_BATTING_HR, 
         'Strikeouts by batters' = TEAM_BATTING_SO,  
         'Homeruns allowed' = TEAM_PITCHING_HR) %>%
  gather("stat", "value") %>%
  filter(complete.cases(.) == TRUE)

@@

* **Homeruns allowed** (average = ~100/year) and **Homeruns by batters** (average = ~106/year) 
have a very similar mid-range distribution (50% of the data lies between ~50 and 150). 
The slight difference in average stats means that teams tend to have a higher number
of homeruns than the opposition team. 
* The only thing that stands out about **Strikeouts by batters** (average = ~736/year) is 
how nearly perfectly normal it is. 50% of the data is between about 500 and 1000 and 
there are absolutely no outliers in the dataset! This means that there were no surprisingly
high or low seasons.

@@

* All of the outliers for **Errors** and **Hits allowed** are above the upper 
tail of the data set. This is further illustrated by the mean and median values
for both of these stats; in both instances, the mean per year (Errors = ~246/year
and Hits allowed = ~1779/year) are higher than the median per year (Errors = 
~159/year and Hits allowed = ~1518/year). This means that some seasons with
exceptionally high values for both of these statistics skew the dataset. 
* The maxiumum value

@@

>>>>

---
title: 'DATA 621---Business Analytics and Data Mining'
subtitle: 'Fall 2020---Group 2---Homework #1'
author: Avraham Adler, Samantha Deokinanan, Amber Ferger, John Kellogg,
    Bryan Persaud, Jeff Shamp
date: "9/27/2020"
output:
  html_document: default
  pdf_document: default
---

@@

```{r loadData, include=FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(ggplot2)
library(corrplot)
library(data.table)
library(tidymodels)
library(mice)
library(psych)
train_set <- read.csv('data/moneyball-training-data.csv')
eval_set <- read.csv('data/moneyball-evaluation-data.csv')

set.seed(9450)
```

@@

We are given a data set of 2,276 records containing 15 seasonal statistics and the
total number of wins a team had in a given year. On average, about 50% of games 
played are won (81 games out of 162), with the best individual season having 146 wins and the worst 
season having 0 wins. The data is normally distributed and most years have between 49 
and 112 wins (blue lines below). The nature of the distribution means there aren't too 
many extreme seasons where wins are significantly higher or lower than usual. This serves 
as a good gut-check for our final predictions; if the predicted wins are too high or too 
low, we know something in our model is probably off.

@@

**Stats with an expected negative impact: ** Intuitively, we expect that Caught 
stealing, Errors, Hits allowed, Home Runs allowed, Strikeouts by batters, and Walks 
allowed would all have a **negative** impact on the total wins. In other words, as 
these values increase, we expect that the team is less likely to win.

@@

negSet <- train_set %>%
  select(TARGET_WINS, 
         'Errors' = TEAM_FIELDING_E, 
         'Hits allowed' = TEAM_PITCHING_H, 
         'Strikeouts by batters' = TEAM_BATTING_SO, 
         'Caught stealing' = TEAM_BASERUN_CS, 
         'Walks allowed' = TEAM_PITCHING_BB, 
         'Home Runs allowed' = TEAM_PITCHING_HR)

@@

Even more interestingly, Home Runs allowed and Walks allowed have the *opposite* 
effect; as these stats increase, so do the number of wins!

**Stats with an expected positive impact: ** We can look at the same information for the 
stats that we expect to have a **positive** effect on wins: Base hits, Doubles, Triples, 
Home Runs, Walks, Batters getting hit by pitches, Stolen bases, Double Plays, and Strikeouts 
by pitchers.

@@

posSet <- train_set %>%
  select(TARGET_WINS, 
         'Base hits' = TEAM_BATTING_H, 
         'Doubles' = TEAM_BATTING_2B, 
         'Triples' = TEAM_BATTING_3B, 
         'Home Runs' = TEAM_BATTING_HR, 
         'Walks' = TEAM_BATTING_BB, 
         'Batters hit' = TEAM_BATTING_HBP,
         'Stolen bases' = TEAM_BASERUN_SB,
         'Double plays' = TEAM_FIELDING_DP,
         'Strikeouts' = TEAM_PITCHING_SO)

@@

Many of these stats *do* seem to have an effect on the number of wins, most notably,
base hits and walks. We see weaker positive relationships for home runs, doubles, triples, 
and stolen bases. This makes sense when we think about it; these things tend to happen 
less often in games than pure base hits and walks, so they don't have as much of an 
effect on winning. Finally, double plays and batters hit don't appear to have any 
correlation with the number of wins. Once again, this intuitively makes sense because 
they are less likely to happen in a game.

@@

Some of the provided statistics are well-balanced in the sense that there are
very few (or no) extreme values. **Home Runs by batters**, **Strikeouts by** 
**batters**, and **Home Runs allowed** are examples of this.

@@

set1 <-  train_set %>%
  select('Home Runs by batters' = TEAM_BATTING_HR, 
         'Strikeouts by batters' = TEAM_BATTING_SO,  
         'Home Runs allowed' = TEAM_PITCHING_HR) %>%
  gather("stat", "value") %>%
  filter(complete.cases(.) == TRUE)

@@

* **Home Runs allowed** (average = ~100/year) and **Home Runs by batters** (average = ~106/year) 
have a very similar mid-range distribution (50% of the data lies between ~50 and 150). 
The slight difference in average stats means that teams tend to have a higher number
of Home Runs than the opposition team. 
* The only thing that stands out about **Strikeouts by batters** (average = ~736/year) is 
how nearly perfectly normal it is. 50% of the data is between about 500 and 1000 and 
there are absolutely no outliers in the dataset! This means that there were no surprisingly
high or low seasons.

@@

* All of the outliers for **Errors** and **Hits allowed** are above the upper 
tail of the data set. This is further illustrated by the mean and median values
for both of these stats; in both instances, the mean per year (Errors = ~246/year
and Hits allowed = ~1779/year) are higher than the median per year (Errors = 
~159/year and Hits allowed = ~1518/year). This means that some seasons with
exceptionally high values for both of these statistics skew the dataset. 
* The maximum value

@@

## Data Preparation

Now we have a good idea of the data we are looking at we can take the next steps to 
prepare it for building a solid model.  We saw from the above that at least one 
variable should be omitted for low reporting of data. There are also outlier concerns 
for several of the variables, namely Team Pitching Strikeouts (`TEAM_PITCHING_SO`), 
Team Batting Hits (`TEAM_BATTING_H`) and Team Pitching Hits ('TEAM_PITCHING_H').

First, we will look into obvious outliers in the data set and remove them. There appears 
to be enough data in the training set to take a naive yet cautious approach to outlier removal.

```{r load, echo=FALSE}
train_set <-
  train_set %>%
  select(-c(TEAM_BATTING_HBP, INDEX)) %>%
  mutate_if(is.integer, as.numeric)
```

### Isolate base hits.

The variable Team Batting Hits (`TEAM_BATTING_H`) is a combination of all hits for 
the season. So we can cleanly use the Second base ('TEAM_BATTING_2B'), Third Base 
('TEAM_BATTING_3B') and Home Run ('TEAM_BATTING_HR') variables we create a First base variable
('TEAM_BATTING_1B') by subtracting the 2B, 3B, and HR respectively from the Team Batting Hits total.

~Do we need a graph here?

```{r singles, echo=FALSE}
train_set<- 
    train_set %>%
    mutate(TEAM_BATTING_1B =
           TEAM_BATTING_H -
           (TEAM_BATTING_2B + TEAM_BATTING_3B + TEAM_BATTING_HR))

### Outlier Removal

When performing the processes of outlier removal a cautious approach is always best.  Each 
outlier is evaluated to ensure its clearly from incorrectly or misentered data, the removal 
does not affect our later assumptions, and the outlier creates a significant association or relation 
at without it the relation is no longer present.

There are some clear issues with two of the variables.  Since both appear with a heavy right skew, we 
will use median and IQR to remove the outliers.

* TEAM_PITCHING_SO
  +We discussed earlier Team Pitching Strikeouts.
* TEAM_PITCHING_H 
  + Team Pitching Hits has a value over 30,000 and two over 20,000
  + These values seem highly unlikely and will be adjusted

```{r outliers_1, echo=FALSE}
train_set %>%
  pivot_longer(everything(), 
               names_to ="Variable", 
               values_to="Value") %>%
  ggplot(aes(x=Variable, y=Value)) +
  geom_boxplot(na.rm = TRUE) + coord_flip()
```

```{r iqr_removal_SO, echo=FALSE}
train_set<-
train_set %>%
  na.omit() %>%
  summarise(iqr = IQR(TEAM_PITCHING_SO)) %>%
  bind_cols(
            train_set 
            )  %>%
  filter(
    TEAM_PITCHING_SO > quantile(TEAM_PITCHING_SO,
                                probs=c(0.25), 
                                na.rm = TRUE) - 1.5*iqr, 
    TEAM_PITCHING_SO < quantile(TEAM_PITCHING_SO, 
                                probs=c(0.75), 
                                na.rm=TRUE) + 1.5*iqr
        ) %>%
  select(-iqr)
```

```{r iqr_removal_H, echo=FALSE}
train_set<-
train_set %>%
  na.omit() %>%
  summarise(iqr = IQR(TEAM_PITCHING_H)) %>%
  bind_cols(
            train_set 
            )  %>%
  filter(
    TEAM_PITCHING_H > quantile(TEAM_PITCHING_H,
                                probs=c(0.25), 
                                na.rm = TRUE) - 1.5*iqr, 
    TEAM_PITCHING_H < quantile(TEAM_PITCHING_H, 
                                probs=c(0.75), 
                                na.rm=TRUE) + 1.5*iqr
        ) %>%
  select(-iqr)

```{r outlier_boxplot, echo=FALSE}
train_set %>%
  pivot_longer(everything(), 
               names_to ="Variable", 
               values_to="Value") %>%
  ggplot(aes(x=Variable, y=Value)) +
  geom_boxplot(na.rm = TRUE) + coord_flip()
```

This has eliminated what appears to be the most extreme outliers. There are 
still large outlier sets for Errors and stolen bases, but none that seem to 
dwarf the other variables. According to the box plot, things appear to be on a 
scale that seems logical given the data and source of information that we have.

## "MICE" imputation method and Feature Engineering

From our early exploration of the data, the vast majority of the data is complete 
with only a few variables with missing values. We will use the "mice" imputation method, 
specifically predictive mean matching method.

MICE (Multivariate imputation by chained equations) is a principled method of dealing 
with missing data.  It creates multiple imputations, as opposed to single imputations 
and accounts for the statistical uncertainty in the imputations.  It creates predictive 
values for the mean instead of imputing the IQR values.

[described here](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3074241/#:~:text=Missing%20data%20are%20a%20common,method%20of%20addressing%20missing%20data.).  
```{r impute_train,echo=FALSE, message=FALSE, warning=FALSE}
train_set <- complete(mice(data = train_set, 
                         method = "pmm", 
                         seed = 9450, 
                         print=FALSE), 3)
```

With outliers removed and missing values imputed via MICE, we can build a feature, 
specifically one for slugging.

### Slugging Percentage

~Why are we building a feature?

Slugging is calculated in the following way

${\displaystyle \mathrm {SLG} ={\frac {({\mathit {1B}})+(2\times {\mathit {2B}})+(3\times {\mathit {3B}})+(4\times {\mathit {HR}})}{AB}}}$

We will take the same approximation for at-bats.

${\displaystyle \mathrm {\sim SLG} ={\frac {({\mathit {1B}})+(2\times {\mathit {2B}})+(3\times {\mathit {3B}})+(4\times {\mathit {HR}})}{SO_{batting} + H + BB}}}$

```{r make_slg_obp, echo=FALSE}
train_set<-
train_set %>%
  mutate(AB = TEAM_BATTING_SO + TEAM_BATTING_BB + TEAM_BATTING_H,
         SLG = (TEAM_BATTING_1B + (2*TEAM_BATTING_2B)+ (3*TEAM_BATTING_3B)+ (4*TEAM_BATTING_HR))/AB) %>%
  # Removing AB and total hits
  select(-c(AB, TEAM_BATTING_H))
```

## Modeling

We will start with a simple linear model and select out features based on null hypothesis \
testing for non-zero slope.

Below is the output of coefficients for backward selection model.

```{r lm_first, echo=FALSE, eval=FALSE}
lm_reg = lm(data=train_set, TARGET_WINS ~ .)
summary(lm_reg)
```

```{r lm_selected, echo=FALSE}
lm_1<- lm(TARGET_WINS ~ TEAM_BATTING_2B +TEAM_BATTING_HR+
          TEAM_BATTING_BB+TEAM_BATTING_SO+ TEAM_BASERUN_SB+TEAM_BASERUN_CS+
          TEAM_PITCHING_HR+ TEAM_PITCHING_H+
          TEAM_FIELDING_E+ TEAM_FIELDING_DP+  SLG, data=train_set)
knitr::kable(lm_1$coefficients, digits = 3L)
```
